Refactoring of Application Using Relational Database

### 2013-02-22 doc/paper.tex
review from David

\subsection{Solution Proposal}
We believe there exists a better solution which automatizes the process of database evolution according to the code refactoring. The solution is illustrated in the Fig. \ref{fig:evolution}. It is based on a change in the evolution process which assumes that the ORM does not change during the evolution:
\begin{enumerate}
	\item The evolution of the whole software is defined independently on the application or the database.
	\item The evolution is interpreted for the application and the database.
	\item The evolution is executed.
\end{enumerate}
The process can decrease the human mistakes in the evolution process, because there is only one source for the evolution and the evolution is automatically interpreted for the application (as a refactoring) and database (as schema and data migration). The existence of the set of all possible software evolutions $E$ is based on a evolutionary transformations specific for application and database. Each transformation contains conditions of transformation feasibility, thus the feasibility of the evolution can be verified.

@@

\subsection{Software Model}
The software consists of three important components (as seen in Fig. \ref{fig:evolution}: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM.
The software has to be in consistent state so its users could have benefit from its usage. The software is in consistent state if application and database are consistent and the database structure corresponds to the application structure according to the ORM, therefore the the software is defined as:
\begin{align}
& software(a, d, \rho) = \begin{cases}
 software(a, d, \rho) \; \textbf{if} \; a \neq \perp \; \wedge \; b \neq \perp \\ \wedge \; \rho(a) = schema(d) 
 \\\\
 \perp
 \end{cases}\\
& a \in Application, d \in Database, \rho \in ORM \nonumber
\end{align}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations.

@@

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. The properties of a class can be of two kinds - values or collections - according of theirs cardinality.

@@

\subsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformation for its creating, altering and removing parts of the model (the set $A$). The conditions for successful transformations are simple in case of creating or altering:
\begin{enumerate}
	\item The name collisions has to be prevent when creating an class in the model or altering its name.
	\item All references have to be updated when renaming a class.
	\item The existence of referenced class has to be verified when creating or altering an association.  
\end{enumerate}
The only issue during deleting classes etc. from an application is the class can be removed only in case it is not associated by other classes. The transformations from $A$ are in table \ref{tab:app-evolution}. The list of altering operations is not complete as there are missing transformation for changing the obligation or cardinality of properties and associations. It is because these transformations are not so important for complex refactorings. 
\begin{table}
\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\centering
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	Application Creation
	& newProperty(className, property, application) \\
	& newAssociation(className, association, mapping, \\ &  \hspace{0.5in}application) \\
	&  newClass(class, application) \\
	Application Alternation
		& renameProperty(className, oldName, newName, \\ &  \hspace{0.5in} application) \\
		& renameAssociation(className, oldName, newName, \\ &  \hspace{0.5in}application) \\
		& renameClass(oldName, newName, application)\\
		& changeReference(referncedClass, newReferencedClass) \\
		Application Disassembly 
		& removeProperty(className, propertyName,\\ &  \hspace{0.5in} application) \\
		& removeAssociation(className, associationName, \\ &  \hspace{0.5in}application) \\
		& removeClass(className, application)\\
%	Copy Structure
%		& copyProperty(sourceClass, property, targetClass,\\ &  \hspace{0.5in} newName, application)  \\
%		& copyClass(class, newName, application) \\
	Inheritance manipulation   & setParent(class, parentClass) \\
		& removeParent(childClass) \\
		& pushDown(class, property)\\
		& pullUp(class, property)\\
	\hline
	\end{tabular}
\end{table}

@@

\subsection{Database Manipulation}
\label{sec:db-evolution}
The database consists of two parts - from database schema which defines the structure and from stored data - hence the transformations from set $D$ have to consider both parts. The transformation for manipulation of structure has similar conditions to the evolution of application application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. The set of transformations for database manipulation is limited in contrast with the SQL language as we mentioned only transformation necessary for data evolution on database level.

@@

\paragraph{Copy Property}
The $copyProperty$ creates a duplicate of a property in a given class. It can rename the property so it can be used for creating duplicates of a property in the property owning class. If the mapping is not provided the transformation creates only a structural copy of the property otherwise it copies the values too. Therefore the  $copyProperty$ transformation is the simplest way to manipulate stored data.
\begin{align}
& copyProperty(sourceClass, propertyName, targetClass, \nonumber  \\
& \; \; \; mapping, software) 
\end{align}
The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \nonumber  \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align}
in case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align}
& \Phi(copyProperty(sourceClass, property, targetClass, newName, \nonumber \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align}
The \emph{copyProperty} transformation gets a mapping between instances (stored data) on imput. The mapping to provides information which value from the source table rows is copied to which row in target table. The mapping is usually defined as: 
\begin{align}\label{eq:mapping}
&	mapping : select(ORM(sourceName), database) \nonumber \\ 
& \;\;\;\;\; \rightarrow  (select(ORM(targetName), database) 
\end{align}
However different definitions can be used the only condition is the result of the mapping has to respect the structure of the instances and constraints on column (e.g. $NULL$ cannot be inserted into a column with $NOTNULL$ constraint defined) The character of transformation depend on the feature of provided mapping. There occurs no data lost if the mapping is surjective.

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copyProperty$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand special attention has to be paid to the provided mapping of instances, because it can cause data loses. The ideal case is when the mapping is surjective, then the transformation can not cause the lost of data. %On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align}
& moveProperty(sourceClass, property, targetClass, newName,\nonumber  \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \nonumber \\
& \;\;\; copyProperty(sourceClass, property, targetClass, \nonumber \\ 
& \;\;\;newName, mapping, software))
\end{align}

@@

Next transformations connected with inheritance are \emph{pushDown} and \emph{pullUp}. Pull up moves a property from child to parent so the \emph{moveProperty} can be used, the mapping of instances is based on the parent-child relationship. It means the column cannot be constrained with $NOTNULL$ constraint if there are more siblings in the hierarchy. The \emph{pushDown} transformation works in opposite direction. The easiest situation is where there no sibling in the hierarchy, then the \emph{moveProperty} can be used, otherwise we assume there are no stored data in the siblings. It is because the change of parent changes affects the instances of its children, when a property is moved only into one children the information consistency is violated, therefore we forbid such transformation, because we cannot anticipate developers intents.

The next two inheritance-related transformations are \emph{extractParent} and \emph{extractCommonParent} serves for extracting a parent class in based on given set of property from a class or from a couple of classes. The transformation is based on the \emph{extractClass} (and \emph{mergeClasses} in the second case).

@@

To improve the design of code the class Address has to be created which associated with both original classes and contains the data already stored in the software. 
\begin{figure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | }
	 	\hline
		Id &  Surname & Street & City & ZIP  \\ \hline  
		11 & Jackson & Central Park St & New York & 100 01  \\ \hline
		12 & Clooney & S Orange Ave & Orlando & 320 24  \\ \hline
	\end{tabular}
	\caption{Data stored in the table $Person$.}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | c |}
	 	\hline
		Id &  BusinessName & Street & City & ZIP \\ \hline  
		100 & Tools \& Machines & Olive ave. & New York & 100 01 \\ \hline
		200 & AI Robotics & Pine ave. & LA & 900 03  \\ \hline
	\end{tabular}
	\caption{Data stored in the table $LegalParty$.}
\end{subfigure}
	\caption{The initial state of the example used in the case study. There is a repetition of information structure in the $Person$ and $LegalParty$ class.}
	\label{fig:case1}
\end{figure}
The first  step of is to extract two temporary classes representing address of a $Person$ or of a $LegalParty$:
\begin{align}
s' &= extract("LegalParty",  "Address\_tmp2", "street", "city", "zip", \varnothing, \nonumber  \\ 
& \;\;\;\; extract("Person", "Address\_tmp1", "street", "city", "zip", \nonumber \\ & \;\;\;\;  \varnothing, software)))
\end{align}
Temporary classes has to be connected with the origin classes by an association. The associations are creates as optional because foreign key constrained with $NOTNULL$ can cause problems in the next phase of evolution.
\begin{align}
 s'' &= newAssociation("LegalParty", association("address", \nonumber \\
& \;\;\;\; "Address\_tmp2", 0, 1), mapping_2, \nonumber \\ & \;\;\;\; newAssociation("Person", association("address",\nonumber  \\
& \;\;\;\; "Address\_tmp1", 0, 1), mapping_1, s')))
\end{align}
The $mapping_1$ and $mapping_2$ can be defined using the equality of primary keys values, because the temporary classes were extracted from the $Person$ and $LegalParty$. Finally the temporary class are merged into one:
\begin{align}
 s''' &= merge("Address\_tmp1", "Address\_tmp2", "Address", s'')
\end{align}
The merge of classes is possible because the associations (and corresponding foreign keys) created in the previous step are not constrained. The result of the transformations is in the Fig. \ref{fig:case2}. 
\begin{figure}
\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\includegraphics[width=\textwidth]{./images/case_app_9}
	\caption{The application model.}
\end{subfigure}
\quad
\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\includegraphics[width=\textwidth]{./images/case_db_8}
	\caption{The database model.}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | l |}
	 	\hline
		Id & Street & City & ZIP & Person\_Address & LegalParty\_Address\\ \hline  
		1 & Olive ave. & New York & 100 01 & & 100 \\ \hline
		2 & Pine ave. & LA & 900 03 & & 200  \\ \hline
		3 & Central Park St & New York & 100 01 & 11 &  \\ \hline
		4 & Orange Ave & Orlando & 320 24 & 12 &\\ \hline
	\end{tabular}
	\caption{Data stored in the table $Address$.}
\end{subfigure}
	\caption{The final state of the example used in the case study. There is only one table containing all addresses in the system, which reference the $Party$ and $LegalParty$.}
	\label{fig:case2}
\end{figure}
The change of the code can continue e.g. by adding more addresses to the $Person$ class, creating an enumeration for cities etc.

The example shows that defined transformations are able to perform regular data evolutions. It shows that usage of the transformations are sometimes not intuitive for user (e.g. creating temporary classes before merge), therefore next complex transformations has to be define based on user experience. The examples shows that the column constraint s often decide the feasibility of a transformation.

@@

The formal definition of MDD approach to database schema evolution is proposed in \cite{Aboulsamh:FormalModelingEvolution}, where a changes of database conceptual schema is interpreted on both physical schema and data. In contrast our proposal is aimed to solve the problem of code refactoring and its impact on relational database, next we propose more complex transformations (such as \emph{copyProperty} or inheritance-related transformations) and examine the impact of platform specific constructs (such as foreign keys or constraints) on the evolution.

>>>>

\subsection{Solution Proposal}
We believe there exists a better solution which automatizes the process of database evolution according to the code refactoring. The solution is illustrated in the Fig.~\ref{fig:evolution}. It is based on a change in the evolution process which assumes that the ORM does not change during the evolution:
\begin{enumerate}
	\item The evolution of the whole software is defined independently on the application or the database.
	\item The evolution is interpreted for the application and the database.
	\item The evolution is executed.
\end{enumerate}
The process can decrease the human mistakes in the evolution process, because there is only one source for the evolution and the evolution is automatically interpreted for the application (as a refactoring) and database (as schema and data migration). The existence of the set of all possible software evolutions $E$ is based on a evolutionary transformations specific for application and database. Each transformation contains conditions of transformation feasibility, thus the feasibility of the evolution can be verified.

@@

\subsection{Software Model}
The software consists of three important components (as seen in Fig. \ref{fig:evolution}: application (its persistent layer concretely), database (consisting of database schema and stored data) and an ORM, therefore we define a software as a triple consisting of an application and a database which are connected together by an ORM.
The software has to be in consistent state so its users could have benefit from its usage. The software is in consistent state if application and database are consistent and the database structure corresponds to the application structure according to the ORM, therefore the software is defined as:
\begin{align}
& software(a, d, \rho) = \begin{cases}
 software(a, d, \rho) \; \textbf{if} \; a \neq \perp \; \wedge \; b \neq \perp \\ \wedge \; \rho(a) = schema(d) 
 \\\\
 \perp
 \end{cases}\\
& a \in Application, d \in Database, \rho \in ORM \nonumber
\end{align}
The evolution of a software is a transformation from one consistent state to another one. These states are called generations of the software and the function which changed the state of a software are called transformations.

@@

\paragraph{Class} Class represents a basic structural unit in the application model. It has a unique name, one or more properties and it can be associated to other classes in the application. The properties of a class can be of two kinds - values or collections - according to theirs cardinality.

@@

\subsection{Application Manipulation}
\label{sec:app-evolution}
The application model defines only the structure of an application thus there are defined transformations for its creating, altering and removing parts of the model (the set $A$). The conditions for successful transformations are simple in case of creating or altering:
\begin{enumerate}
	\item The name collisions has to be prevented when creating an class in the model or altering its name.
	\item All references have to be updated when renaming a class.
	\item The existence of referenced class has to be verified when creating or altering an association.  
\end{enumerate}
The only issue during deleting classes etc. from an application is the class can be removed only in case it is not associated by other classes. The transformations from $A$ are in table \ref{tab:app-evolution}. The list of altering operations is not complete as there are missing transformation for changing the obligation or cardinality of properties and associations. It is because these transformations are not so important for complex refactorings. 
\begin{table}
\caption{The transformations for application evolution.}
	\label{tab:app-evolution}
\centering
	\begin{tabular}{ll}
	\hline
	Type of transformation & Defined transformations \\
	\hline
	\\[-2ex] 
	Application Creation
	& newProperty(className, property, application) \\
	& newAssociation(className, association, mapping, \\ &  \hspace{0.5in}application) \\
	&  newClass(class, application) \\
	Application Alternation
		& renameProperty(className, oldName, newName, \\ &  \hspace{0.5in} application) \\
		& renameAssociation(className, oldName, newName, \\ &  \hspace{0.5in}application) \\
		& renameClass(oldName, newName, application)\\
		& changeReference(referncedClass, newReferencedClass) \\
		Application Disassembly 
		& removeProperty(className, propertyName,\\ &  \hspace{0.5in} application) \\
		& removeAssociation(className, associationName, \\ &  \hspace{0.5in}application) \\
		& removeClass(className, application)\\
%	Copy Structure
%		& copyProperty(sourceClass, property, targetClass,\\ &  \hspace{0.5in} newName, application)  \\
%		& copyClass(class, newName, application) \\
	Inheritance manipulation   & setParent(class, parentClass) \\
		& removeParent(childClass) \\
		& pushDown(class, property)\\
		& pullUp(class, property)\\
	\hline
	\end{tabular}
\end{table}

@@

\subsection{Database Manipulation}
\label{sec:db-evolution}
The database consists of two parts - from database schema which defines the structure and from stored data - hence the transformations from set $D$ have to consider both parts. The transformation for manipulation of structure has similar conditions to the evolution of application. The transformations for data manipulation are inspired by the SQL language. The basic transformations are in Table \ref{tab:db-basic-evolution}. The set of transformations for database manipulation is limited in contrast with the SQL language as we mentioned only transformation necessary for data evolution on database level.

@@

\paragraph{Copy Property}
The $copyProperty$ creates a duplicate of a property in a given class. It can rename the property so it can be used for creating duplicates of a property in the property owning class. If the mapping is not provided the transformation creates only a structural copy of the property otherwise it copies the values too. Therefore the  $copyProperty$ transformation is the simplest way to manipulate stored data.
\begin{align}
& copyProperty(sourceClass, propertyName, targetClass, \nonumber  \\
& \; \; \; mapping, software) 
\end{align}
The  $copyProperty$ is the first transformation when an additional information has to be added to the usual code refactoring. It is because the $copyColumn$ transformation needs one more information to succeed - there has to be provided mapping between the source and target table to assure the data information consistency. %Moreover the source and target should have the same structure, type and constraints.
The transformation is interpreted for both software components - in case of the application a new property is added:
\begin{align}
& \Psi(copyProperty(sourceClass, property, targetClass, mapping, \nonumber  \\ 
&  \; \; \; software)) = newProperty(targetClass, property, application) 
\end{align}
in case of database the copy of a column or table is created according to property's cardinality and then values are copied:
\begin{align}
& \Phi(copyProperty(sourceClass, property, targetClass, newName, \nonumber \\ 
&  \; \; \; mapping, software)) = \begin{cases}
 coypColumn(ORM(sourceClass),  \\ \;\;\; ORM(targetClass),  \\ \;\;\;alterColumnName(ORM(property), \\ \;\;\; ORM(newName)), mapping, database) \\ \; \mathbf{if} \; cardinality(property) \leq 1 \\\\
 copyTable(ORM(Property), \\ \;\;\; ORM(newName))
 \end{cases}
\end{align}
The \emph{copyProperty} transformation gets a mapping between instances (stored data) on input. The mapping to provides information which value from the source table rows is copied to which row in target table. The mapping is usually defined as: 
\begin{align}\label{eq:mapping}
&	mapping : select(ORM(sourceName), database) \nonumber \\ 
& \;\;\;\;\; \rightarrow  (select(ORM(targetName), database) 
\end{align}
However different definitions can be used the only condition is the result of the mapping has to respect the structure of the instances and constraints on column (e.g. $NULL$ cannot be inserted into a column with $NOTNULL$ constraint defined) The character of transformation depends on the feature of provided mapping. There occurs no data lost if the mapping is surjective.

\paragraph{Move Property}
The $moveProperty$ transformation is based on the $copyProperty$ transformation followed by the $removeProperty$ so its construction is easy. On the other hand special attention has to be paid to the provided mapping of instances, because it can cause lost of data. The ideal case is when the mapping is surjective, then the transformation can not cause the lost of data. %On the other hand the situation when the mapping does not provides pairs for all rows in a source table can lead to data loss. 
\begin{align}
& moveProperty(sourceClass, property, targetClass, newName,\nonumber  \\
& \;\;\; mapping, software) = removeProperty(sourceClass, property, \nonumber \\
& \;\;\; copyProperty(sourceClass, property, targetClass, \nonumber \\ 
& \;\;\;newName, mapping, software))
\end{align}

@@

Next transformations connected with inheritance are \emph{pushDown} and \emph{pullUp}. Pull up moves a property from child to parent so the \emph{moveProperty} can be used, the mapping of instances is based on the parent-child relationship. It means the column cannot be constrained with $NOTNULL$ constraint if there are more siblings in the hierarchy. The \emph{pushDown} transformation works in opposite direction. The easiest situation is where there is no sibling in the hierarchy, then the \emph{moveProperty} can be used, otherwise we assume there are no stored data in the siblings. It is because the change of parent affects the instances of its children, when a property is moved only into one children the information consistency is violated, therefore we forbid such transformation, because we cannot anticipate developers intents.

The next two inheritance-related transformations are \emph{extractParent} and \emph{extractCommonParent} which serve for extracting a parent class based on given set of property from a class or from a couple of classes. The transformation is based on the \emph{extractClass} (and \emph{mergeClasses} in the second case).

@@

To improve the design of code the class Address has to be created which associated with both original classes and contains the data already stored in the software. 
\begin{figure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | }
	 	\hline
		Id &  Surname & Street & City & ZIP  \\ \hline  
		11 & Jackson & Central Park St & New York & 100 01  \\ \hline
		12 & Clooney & S Orange Ave & Orlando & 320 24  \\ \hline
	\end{tabular}
	\caption{Data stored in the table $Person$.}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | c |}
	 	\hline
		Id &  BusinessName & Street & City & ZIP \\ \hline  
		100 & Tools \& Machines & Olive ave. & New York & 100 01 \\ \hline
		200 & AI Robotics & Pine ave. & LA & 900 03  \\ \hline
	\end{tabular}
	\caption{Data stored in the table $LegalParty$.}
\end{subfigure}
	\caption{The initial state of the example used in the case study. There is a repetition of information structure in the $Person$ and $LegalParty$ class.}
	\label{fig:case1}
\end{figure}
The first  step is to extract two temporary classes representing address of a $Person$ or of a $LegalParty$:
\begin{align}
s' &= extract("LegalParty",  "Address\_tmp2", "street", "city", "zip", \varnothing, \nonumber  \\ 
& \;\;\;\; extract("Person", "Address\_tmp1", "street", "city", "zip", \nonumber \\ & \;\;\;\;  \varnothing, software)))
\end{align}
Temporary classes has to be connected with the origin classes by an association. The associations are created as optional because foreign key constrained with $NOTNULL$ can cause problems in the next phase of evolution.
\begin{align}
 s'' &= newAssociation("LegalParty", association("address", \nonumber \\
& \;\;\;\; "Address\_tmp2", 0, 1), mapping_2, \nonumber \\ & \;\;\;\; newAssociation("Person", association("address",\nonumber  \\
& \;\;\;\; "Address\_tmp1", 0, 1), mapping_1, s')))
\end{align}
The $mapping_1$ and $mapping_2$ can be defined using the equality of primary keys values, because the temporary classes were extracted from the $Person$ and $LegalParty$. Finally the temporary class are merged into one:
\begin{align}
 s''' &= merge("Address\_tmp1", "Address\_tmp2", "Address", s'')
\end{align}
The merge of classes is possible because the associations (and corresponding foreign keys) created in the previous step are not constrained. The result of the transformations is in the Fig. \ref{fig:case2}. 
\begin{figure}
\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\includegraphics[width=\textwidth]{./images/case_app_9}
	\caption{The application model.}
\end{subfigure}
\quad
\begin{subfigure}[b]{0.45\textwidth}
	\centering
	\includegraphics[width=\textwidth]{./images/case_db_8}
	\caption{The database model.}
\end{subfigure}
\begin{subfigure}[b]{\textwidth}
	\centering
	\begin{tabular}{| c | l | l | l | l | l |}
	 	\hline
		Id & Street & City & ZIP & Person\_Address & LegalParty\_Address\\ \hline  
		1 & Olive ave. & New York & 100 01 & & 100 \\ \hline
		2 & Pine ave. & LA & 900 03 & & 200  \\ \hline
		3 & Central Park St & New York & 100 01 & 11 &  \\ \hline
		4 & Orange Ave & Orlando & 320 24 & 12 &\\ \hline
	\end{tabular}
	\caption{Data stored in the table $Address$.}
\end{subfigure}
	\caption{The final state of the example used in the case study. There is only one table containing all addresses in the system, which reference the $Party$ and $LegalParty$.}
	\label{fig:case2}
\end{figure}
The change of the code can continue e.g. by adding more addresses to the $Person$ class, creating an enumeration for cities etc.

The example shows that defined transformations are able to perform regular data evolutions. It shows that usage of the transformations are sometimes not intuitive for user (e.g. creating temporary classes before merge), therefore next complex transformations has to be defined based on user experience. The examples shows that the column constraint s often decide the feasibility of a transformation.

@@

The formal definition of MDD approach to database schema evolution is proposed in \cite{Aboulsamh:FormalModelingEvolution}, where changes of database conceptual schema is interpreted on both physical schema and data. In contrast our proposal is aimed to solve the problem of code refactoring and its impact on relational database, next we propose more complex transformations (such as \emph{copyProperty} or inheritance-related transformations) and examine the impact of platform specific constructs (such as foreign keys or constraints) on the evolution.
